Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-16 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
CH 8 and spell check

\acknowledgments{
I would like to thank Prof.~Fleischmann, my senior advisor with the University of Canterbury, Dr.~Kevin McGrattan from the National Institute of Standards and Technology, and Dr.~Craig Weinschenk from the UL Firefigther Safety Research Institute for their guidance, assistance, and support throughout this study.

@@

A series of experiments were conducted with samples of painted gypsum wallboard to determine the repeatability of a ``critical heat flux'' that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel, ASTM~E648, and a modified radiant panel apparatus based on ASTM~E162 were used to examine this value~\cite{ASTM_E648},~\cite{ASTM_E162}.

@@

Cone calorimeter experiments were conducted, based on ASTM~E1354, to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

N.B. This section may be considered atypical for an engineering thesis but it is included here for two reasons.  Firstly, to provide a narrative description for readers that may be unfamiliar with how fire patterns are routinely relied on in fire investigations.  The second is to put this work into an applied context that is familiar to fire investigators that may be un accustomed to reading detailed research as found in this thesis.

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such that the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene; ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On ``side D'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room there is a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less thermal damage you find.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

\subsection{Natural Gas Burner Plume and Wall Exposures}
Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhpas as a result of increased convection on the face of the wall.

@@

A summary of the peak average temperatures and heat fluxes at the non-combustible wall is provided in Table~\ref{tab:TW_Summary}.   Based on the temperature criteria for igniting gyspum wallboard, found in the literature, only the gasoline and polyurethane foam burners positioned against the wall provided temperatures in excess of 515~$^{\circ}$C.  Based on heat flux, each of the burners exceeded the 26~kW/m$^2$ ignition criteria, when positioned against the wall and at when the edge of the burner was 0.15~m away from the wall.

@@

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are simlar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m amd 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented gypsum wall board will provide data for comparsion with data provided in this chapter and provide insight in to where the gypsum wall board may ignite.

@@

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of either its thermal properties or in terms of its physical change as a result of exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, a process commonly known as calcination.  Schroder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction~\cite{Schroeder:1999}.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dihydrated gypsum to map fire patterns.

@@

A complete set of temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set}, \ref{IWGB_NG_TC_Wall_set}, \ref{IWGB_NG_TC_Backside_set}, \ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed, at which point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provided distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature, see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant or continuous flame region appeared to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seemed to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall, the temperatures started at ambient and slowly increased, reaching the peak temperatures at 300~seconds, then the burner was shut off.  For the case where the burner was in contact with the wall, the two thermocouples located in the persistant flame region indicated a decreasing temperature trend after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

  \begin{figure}[ht!]
 	\centering
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG01_HF_Offset_Avg}\\
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG04_HF_Center_Avg}\\
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG05_HF_Offset_Avg}\\
 \caption[Wall heat flux along the endge of the natural gas fueled burner]{Wall heat flux along the vertical edge of the natural gas fueled burner. Each graph displays data from a different distance between the edge of the burner and the instrumented gypsum wallboard.  Beginning with the top graph and moving down, the distances from the wall are 0.31~m, 0.15~m, and 0~m.}
 	\label{IWGB_NG_HF_EdgeSet}
 \end{figure}

@@

With this set of experiments, fire patterns as the result of thermal damage were finally observed in a few cases.  To recap the crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter 3, the gasoline had the potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~m and 0.6~m.  With the polyurethane burner, three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the crude predictions from Chapter 3 provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption calorimetry method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a conical shaped heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

@@

The radiant flooring panel test method is known as ASTM~E648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}.  The radiant flooring panel is used to measure the flame spread of a material while exposed to an incident radiant heat flux. The source of the heat flux is a pre-mixed air-gas-fueled radiant panel.  The radiating surface measures 305~mm by 457~mm. The radiating surface is inclined at 30~$^{\circ}$ above the horizontal. The test specimen is mounted horizontally beneath the radiant panel.  The surface of the test sample exposed to the radiant panel is 1~m long and 0.2~m wide. Under standard operating conditions the radiant panel provides a heat flux distribution on the test specimen varying from 10~kW/m$^2$ at the measurement point, centered 100~mm from the edge of the sample closest to the panel to 1~kW/m$^2$ at the farthest measurement point 900~mm from the edge of the sample closest to the panel. A schematic of the radiant flooring panel apparatus is shown in Figure~\ref{FLR_Radpan}.

@@

A diffusion, propane-fueled, pilot flame from a 254~mm long wide stainless steel burner tube, as specified in ASTM~E648, was used to ignite the test sample near the edge of the sample closet to the panel~\cite{ASTM_E648}. The objective of these experiments was to determine the critical radiant flux for the painted gypsum wallboard samples.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

@@

The vertical radiant panel used is documented in ASTM~E162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that has a radiating surface 305~mm wide by 457~mm high.

@@

\begin{figure}
	\centering
	\includegraphics[width=2.92in]{../Figures/RAD_pan}
	\includegraphics[width=2.08in]{../Figures/RAD_pan_iso} \\
	\caption[Photographs of the radiant panel experimental arrangement]{Photographs of the radiant panel experimental arrangement, on the left is a photograph which shows the radiant panel on the right, the aluminum heat shield in the center and the moveable sample mounting frame trolley for the gypsum wallboard sample on the left.  The photograph on the right shows an isometric view of the test arrangement.}
	\label{Rad_Panel}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/FRP_flamefront_arrows} \\
	\caption[Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel] {Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel. Thermocouples imbedded at 100~mm intervals in the surface of the painted paper can be seen in front of and behind the flame front. The flame is leaning back in the direction the radiant panel.}
	\label{FRP_flamefront_arrows}
\end{figure}

@@

\begin{table}
	\centering
	\small
	\begin{tabular}{|c|c|c|c|c|}
		\hline Postion & Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (mm) & (kW/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\ 
		\hline 100 	& 21	& Yes 	& 133	& 315 	 \\
		\hline 200	& 18 	& Yes 	& 198 	& 296 	 \\
		\hline 300	& 14	& Yes 	& 302 	& 279 	 \\
		\hline 400	& 10	& Yes 	& 502 	& 242 	 \\
		\hline 500	&  7	& No 	& NA 	& NA   	 \\
		\hline
	\end{tabular}
	\caption[Flooring radiant panel experiment flame spread results]{Flooring radiant panel experiment flame spread results including time to ignition at 100~mm intervals and surface temperature just prior to ignition.  The values are averages from three replicate experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for heat flux.}
	\label{tab:FRP_timetemp}
\end{table}

@@

Figure~\ref{NG_Corner} shows a comparison of the ten fire patterns generated by the natural gas burner in the corner position with walls of open back construction. In the corner burns there is an area in the corner where the paper is not burned off.  From the flame sequence shown in Figure~\ref{GBNG34Sequence} it can be seen that the flame in the persitent flame region is close to the walls of the corner, but moving up into the intermittent region, the flame burns away from the walls, hence less energy is transferred to the walls and perhaps a bit more soot condenses out on the cooler walls.

>>>>

\acknowledgments{
I would like to thank Prof.~Fleischmann, my senior advisor with the University of Canterbury, Dr.~Kevin McGrattan from the National Institute of Standards and Technology, and Dr.~Craig Weinschenk from the UL Firefighter Safety Research Institute for their guidance, assistance, and support throughout this study.

@@

A series of experiments were conducted with samples of painted gypsum wallboard to determine the repeatability of a ``critical heat flux'' that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel, ASTM~E 648, and a modified radiant panel apparatus based on ASTM~E 162 were used to examine this value~\cite{ASTM_E648},~\cite{ASTM_E162}.

@@

Cone calorimeter experiments were conducted, based on ASTM~E 1354, to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

N.B. This section may be considered atypical for an engineering thesis but it is included here for two reasons.  Firstly, to provide a narrative description for readers that may be unfamiliar with how fire patterns are routinely relied on in fire investigations.  The second is to put this work into an applied context that is familiar to fire investigators that may be unaccustomed to reading detailed research as found in this thesis.

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older town homes.  The socio-economics of the neighborhood are such that the people that live there have many life struggles. En route you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene; ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On ``side D'' of the town home there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the town home and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room there is a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less thermal damage you find.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 mega-pixel resolution.  A mini DV camcorder with 0.7 mega-pixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

\subsection{Natural Gas Burner Plume and Wall Exposures}
Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhaps as a result of increased convection on the face of the wall.

@@

A summary of the peak average temperatures and heat fluxes at the non-combustible wall is provided in Table~\ref{tab:TW_Summary}.   Based on the temperature criteria for igniting gypsum wallboard, found in the literature, only the gasoline and polyurethane foam burners positioned against the wall provided temperatures in excess of 515~$^{\circ}$C.  Based on heat flux, each of the burners exceeded the 26~kW/m$^2$ ignition criteria, when positioned against the wall and at when the edge of the burner was 0.15~m away from the wall.

@@

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are similar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m and 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented gypsum wall board will provide data for comparison with data provided in this chapter and provide insight in to where the gypsum wall board may ignite.

@@

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of either its thermal properties or in terms of its physical change as a result of exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, a process commonly known as calcination.  Schroeder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction~\cite{Schroeder:1999}.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dehydrated gypsum to map fire patterns.

@@

A complete set of temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set}, \ref{IWGB_NG_TC_Wall_set}, \ref{IWGB_NG_TC_Backside_set}, \ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This discrepancy was not noticed until the data was reviewed, at which point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provided distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature, see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistent or continuous flame region appeared to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seemed to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall, the temperatures started at ambient and slowly increased, reaching the peak temperatures at 300~seconds, then the burner was shut off.  For the case where the burner was in contact with the wall, the two thermocouples located in the persistent flame region indicated a decreasing temperature trend after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gypsum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

  \begin{figure}[ht!]
 	\centering
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG01_HF_Offset_Avg}\\
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG04_HF_Center_Avg}\\
 	\includegraphics[width=.625\columnwidth]{../Figures/IWGBNG05_HF_Offset_Avg}\\
 \caption[Wall heat flux along the edge of the natural gas fueled burner]{Wall heat flux along the vertical edge of the natural gas fueled burner. Each graph displays data from a different distance between the edge of the burner and the instrumented gypsum wallboard.  Beginning with the top graph and moving down, the distances from the wall are 0.31~m, 0.15~m, and 0~m.}
 	\label{IWGB_NG_HF_EdgeSet}
 \end{figure}

@@

With this set of experiments, fire patterns as the result of thermal damage were finally observed in a few cases.  To recap the crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter 3, the gasoline had the potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~m and 0.6~m.  With the polyurethane burner, three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatability of the burn patterns has not been assessed yet, but it appears that the crude predictions from Chapter 3 provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption calorimetry method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E 1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a conical shaped heater, spark ignition, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

@@

The radiant flooring panel test method is known as ASTM~E 648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}.  The radiant flooring panel is used to measure the flame spread of a material while exposed to an incident radiant heat flux. The source of the heat flux is a pre-mixed air-gas-fueled radiant panel.  The radiating surface measures 305~mm by 457~mm. The radiating surface is inclined at 30~$^{\circ}$ above the horizontal. The test specimen is mounted horizontally beneath the radiant panel.  The surface of the test sample exposed to the radiant panel is 1~m long and 0.2~m wide. Under standard operating conditions the radiant panel provides a heat flux distribution on the test specimen varying from 10~kW/m$^2$ at the measurement point, centered 100~mm from the edge of the sample closest to the panel to 1~kW/m$^2$ at the farthest measurement point 900~mm from the edge of the sample closest to the panel. A schematic of the radiant flooring panel apparatus is shown in Figure~\ref{FLR_Radpan}.

@@

A diffusion, propane-fueled, pilot flame from a 254~mm long wide stainless steel burner tube, as specified in ASTM~E 648, was used to ignite the test sample near the edge of the sample closet to the panel~\cite{ASTM_E648}. The objective of these experiments was to determine the critical radiant flux for the painted gypsum wallboard samples.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

@@

The vertical radiant panel used is documented in ASTM~E 162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that has a radiating surface 305~mm wide by 457~mm high.

@@

\begin{figure}
	\centering
	\includegraphics[width=2.92in]{../Figures/RAD_pan}
	\includegraphics[width=2.08in]{../Figures/RAD_pan_iso} \\
	\caption[Photographs of the radiant panel experimental arrangement]{Photographs of the radiant panel experimental arrangement, on the left is a photograph which shows the radiant panel on the right, the aluminum heat shield in the center and the movable sample mounting frame trolley for the gypsum wallboard sample on the left.  The photograph on the right shows an isometric view of the test arrangement.}
	\label{Rad_Panel}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/FRP_flamefront_arrows} \\
	\caption[Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel] {Photograph of the flame front burning on the gypsum wallboard sample in the flooring radiant panel. Thermocouples embedded at 100~mm intervals in the surface of the painted paper can be seen in front of and behind the flame front. The flame is leaning back in the direction the radiant panel.}
	\label{FRP_flamefront_arrows}
\end{figure}

@@

\begin{table}
	\centering
	\small
	\begin{tabular}{|c|c|c|c|c|}
		\hline Position & Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (mm) & (kW/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\ 
		\hline 100 	& 21	& Yes 	& 133	& 315 	 \\
		\hline 200	& 18 	& Yes 	& 198 	& 296 	 \\
		\hline 300	& 14	& Yes 	& 302 	& 279 	 \\
		\hline 400	& 10	& Yes 	& 502 	& 242 	 \\
		\hline 500	&  7	& No 	& NA 	& NA   	 \\
		\hline
	\end{tabular}
	\caption[Flooring radiant panel experiment flame spread results]{Flooring radiant panel experiment flame spread results including time to ignition at 100~mm intervals and surface temperature just prior to ignition.  The values are averages from three replicate experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for heat flux.}
	\label{tab:FRP_timetemp}
\end{table}

@@

Figure~\ref{NG_Corner} shows a comparison of the ten fire patterns generated by the natural gas burner in the corner position with walls of open back construction. In the corner burns there is an area in the corner where the paper is not burned off.  From the flame sequence shown in Figure~\ref{GBNG34Sequence} it can be seen that the flame in the persistent flame region is close to the walls of the corner, but moving up into the intermittent region, the flame burns away from the walls, hence less energy is transferred to the walls and perhaps a bit more soot condenses out on the cooler walls.
